Extended Data

### 2022-02-22 doc/manuscript/Dow_manuscript_Nature.Rmd
clarified conifer references
#115

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021]. 
The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing of growth remains unclear.

>>>>

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**

@@

<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current models assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021] (but see Refs. [@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers). 
The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021; @delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021]. <!--this sentence is repetitive with with previous.-->
